Induced universal graphs for families of small graphs

### 2021-09-02 paper/paper.tex
Brendan McKay's suggestions

The values $f(1)$, $f(2)$ and $f(3)$ are equal to to the simple lower bound $2k
- 1$ given in a question by ``Chain Markov'' on Mathematics Stack Exchange
  \cite{math_se_question}.  The values $f(4)$ and $f(5)$ are equal to a lower
  bound given in a comment by ``bof'' on the same question: $f(k) \geq 2k$ if $k
  \geq 4$.  (For $k < 10$, this bound improves upon Moon's lower bound $f(k)
  \leq 2^{(k-1)/2}$.) To briefly summarise the proof, if $k \leq 2k$ then $G$
  must be a split graph (that is, a graph whose vertices can be partitioned
  into a clique and an independent set); therefore, $G$ cannot contain the
  cycle $C_4$ as an induced subgraph.  An example of an 8-vertex induced
  universal graph for the family of 4-vertex graphs was given by ``Chain
  Markov'' as a comment on the same question.

@@

To give an upper bound of 14 on $f(6)$, Figure~\ref{fig:adjmat14} shows the
adjacency matrix of a 14-vertex graph that is induced universal for the family
of all graphs on six vertices.  This was generated using a simple local search
algorithm.  We begin by generating a random graph on 14 vertices as follows.
We number the vertices from zero; the generator makes the
$k$ vertices numbered $0$ to $k-1$ a clique, and the $k$ vertices numbered $k-1$ to $2k-2$ an
independent set.  Each possible edge that is not involved in either the clique
or the independent set is added with probability $1/2$.
We then repeatedly ``flip'' the status of a random edge from present to absent
or vice versa, but always leave the large clique and independent set intact.
After each flip, we count the number of 156 graphs on 6 vertices that are isomorphic
to a subgraph of our 14-vertex graph.  If the most recent flip resulted in a decrease
in this number, we revert it.  After each 1000 flips, we restart the algorithm
with a new random graph.

@@

By Proposition~\ref{f6proposition}, we have $f(7) \geq 16$.  Figure~\ref{fig:adjmat18}
shows the adjacency matrix of an 18-vertex induced universal
graph for the family of all seven-vertex graphs. This was generated with 
the heuristic described in Section~\ref{sec:f6}, with 10000 rather than 1000
edge-flips permitted before each restart.

@@

>>>>

The values $f(1)$, $f(2)$ and $f(3)$ are equal to to the simple lower bound $2k
- 1$ given in a question by ``Chain Markov'' on Mathematics Stack Exchange
  \cite{math_se_question}.  The values $f(4)$ and $f(5)$ are equal to a lower
  bound given in a comment by ``bof'' on the same question: $f(k) \geq 2k$ if $k
  \geq 4$.  (For $k < 10$, this bound improves upon Moon's lower bound $f(k)
  \leq 2^{(k-1)/2}$.) To briefly summarise the proof, if $f(k) \leq 2k$ then $G$
  must be a split graph (that is, a graph whose vertices can be partitioned
  into a clique and an independent set); therefore, $G$ cannot contain the
  cycle $C_4$ as an induced subgraph.  An example of an 8-vertex induced
  universal graph for the family of 4-vertex graphs was given by ``Chain
  Markov'' as a comment on the same question.

@@

To give an upper bound of 14 on $f(6)$, Figure~\ref{fig:adjmat14} shows the
adjacency matrix of a 14-vertex graph that is induced universal for the family
of all graphs on six vertices.  This was generated using a simple local search
algorithm.  We begin by generating a random graph on 14 vertices as follows.
We number the vertices from zero; the generator makes the
$k$ vertices numbered $0$ to $k-1$ a clique, and the $k$ vertices numbered $k-1$ to $2k-2$ an
independent set.\footnote{The clique and the independent set thus have one vertex
in common.  The proof of Proposition~\ref{f6proposition} can be modified straightforwardly
to show that there is no 14-vertex induced universal graph for this family of graphs
that contains a 6-vertex clique and a 6-vertex independent set as
induced subgraphs with disjoint vertex sets.}
Each possible edge that is not involved in either the clique
or the independent set is added with probability $1/2$.
We then repeatedly ``flip'' the status of a random edge from present to absent
or vice versa, but always leave the large clique and independent set intact.
After each flip, we count the number of 156 graphs on 6 vertices that are isomorphic
to a subgraph of our 14-vertex graph.  If the most recent flip decreased
this number, we revert it.  After each 1000 flips, we restart the algorithm
with a new random graph.

@@

By Proposition~\ref{f6proposition}, we have $f(7) \geq 16$.  Figure~\ref{fig:adjmat18}
shows the adjacency matrix of an 18-vertex induced universal
graph for the family of all seven-vertex graphs. This was generated with 
the heuristic described in Section~\ref{sec:f6}, with two modifications.
First, 10000 rather than 1000
edge-flips were permitted before each restart, as this was found to be more effective
in a preliminary run of the experiment.  Second, the overlapping six-vertex clique
and independent set were replaced with a clique on seven vertices and an independent
set on seven vertices.  Again, these had one vertex in common.  (We also tried making
the clique and independent set vertex-disjoint, but did not find an 18-vertex solution
in four hours with this approach.)

@@

\section{Acknowledgements}

I would like to thank Brendan McKay for helpful feedback on this paper, and Persi
Diaconis for introducing me to induced universal graphs and for interesting email
discussions on related topics.
